---
title: "Task 2"
author: "Max Settineri"
date: "2023-02-20"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(broom)
library(patchwork)
library(janitor)
library(kableExtra)
```

### Overview



**Data Citation:** Lightfoot, D. and W.G. Whitford. 2020. Lizard pitfall trap data from 11 NPP study locations at the Jornada Basin LTER site, 1989-2006 ver 37. Environmental Data Initiative. https://doi.org/10.6073/pasta/4a6e258fb49c31e222ecbbcfd128967f

### Reading in and cleaning data

```{r}
lizards <- read_csv(here("data", "lizards.csv")) %>% 
  clean_names() %>% 
  mutate(log_weight = log(weight)) %>% # log transforming the data - since we are using an exponential base equation
  mutate(log_length = log(sv_length)) %>% 
  drop_na()
```

### Fitting snout length to weight model

**Equation**
\[W = a(SVL)^b\]

- Where `W` is the Body weight (g), `SVL` is snout to vent length (mm).
- `a` and `b` are parameters which need to be fitted


```{r}
# running a guess model
guess_model <- lm(log_weight ~ log_length, data = lizards) 
```

ulling out the coefficients from my_guess_model

- Using `coefficients` function
```{r}
coef <- coefficients(my_guess_model)
#Cleaning up the model so I can pull out the coefficients easier 
my_guess_model_clean <- tidy(my_guess_model)
```
**Coefficient Output**

- The intercept of `my_guess_model` is `r round(my_guess_model_clean$estimate[1],3)`
- The log_length of `my_guess_model` is `r round(my_guess_model_clean$estimate[2],3)`