---
title: "Task 1"
author: "Max Settineri"
date: "2023-02-15"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(broom)
library(patchwork)
library(GGally)
library(ggbeeswarm)
library(tidymodels)
library(AICcmodavg)
library(kableExtra)
```

### Overview

This report....

Questions answered:

- 
- 

**Data source:** Abrahamson, W.G. 2019. Survival, growth and biomass estimates of two dominant palmetto species of south-central Florida from 1981 - 2017, ongoing at 5-year intervals ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/f2f96ec76fbbd4b9db431c79a770c4d5

### Reading in and cleaning up the data

```{r}
palmetto <- read_csv(here('data', 'palmetto.csv')) %>% 
  select(height, length, width, green_lvs, species) %>% # selecting variables of interest
  drop_na() %>% # dropping NA values
  mutate(species_name = case_when(
    species == 1 ~ 'Serenoa repens', # new column for species name with full species names
    species == 2 ~ 'Sabal etonia')) %>% 
    mutate(species_name = as_factor(species_name)) # changing species name to a factor for blr
```

### Exploring differences between Serenoa repens and Sabal etonia

```{r, eval = FALSE}
# using GGally to look at trends and relationships between the two species and four variables of interest
palmetto %>% 
  select(species_name, height:green_lvs) %>% 
  ggpairs(aes(color = species_name))

# after initial analysis, looks to be: strong variation in green_lvs by species, slight variation in length and height, minimal difference in width
```

```{r}
# plotting height by species
plot_h <- ggplot(data = palmetto, aes(x = species_name, y = height)) +
  geom_beeswarm(aes(color = species_name)) +
  theme_minimal() +
  labs(x = "Species", y = "Canopy height (cm)") +
  guides(fill = guide_legend(title = "Species")) +
  geom_boxplot(fill = NA, width = 0.4, outlier.color = NA) +
  theme(axis.ticks.x = element_blank(),
    axis.text.x = element_blank()) +
  scale_color_manual(values = c("cyan4", "orange3")) +
  theme(legend.position = c(0.9, 0.8)) +
  theme(legend.background = element_rect(fill="white", 
                                  size=0, linetype="solid"))
```

```{r}
# plotting length by species
plot_l <- ggplot(data = palmetto, aes(x = species_name, y = length)) +
  geom_beeswarm(aes(color = species_name)) +
  theme_minimal() +
  labs(x = "Species", y = "Length (cm)") +
  guides(fill = guide_legend(title = "Species")) +
  geom_boxplot(fill = NA, width = 0.4, outlier.color = NA) +
  theme(axis.ticks.x = element_blank(),
    axis.text.x = element_blank()) +
  scale_color_manual(values = c("cyan4", "orange3")) +
  theme(legend.position = c(0.9, 0.8)) +
  theme(legend.background = element_rect(fill="white", 
                                  size=0, linetype="solid"))
```

```{r}
# plotting width by species
plot_w <- ggplot(data = palmetto, aes(x = species_name, y = width)) +
  geom_beeswarm(aes(color = species_name)) +
  theme_minimal() +
  labs(x = "Species", y = "Width (cm)") +
  guides(fill = guide_legend(title = "Species")) +
  geom_boxplot(fill = NA, width = 0.4, outlier.color = NA) +
  theme(axis.ticks.x = element_blank(),
    axis.text.x = element_blank()) +
  scale_color_manual(values = c("cyan4", "orange3")) +
  theme(legend.position = c(0.9, 0.8)) +
  theme(legend.background = element_rect(fill="white", 
                                  size=0, linetype="solid"))
```

```{r}
# plotting green leaves by species
plot_gl <- ggplot(data = palmetto, aes(x = species_name, y = green_lvs)) +
  geom_beeswarm(aes(color = species_name)) +
  theme_minimal() +
  labs(x = "Species", y = "Number of green leaves") +
  guides(fill = guide_legend(title = "Species")) +
  geom_boxplot(fill = NA, width = 0.4, outlier.color = NA) +
  theme(axis.ticks.x = element_blank(),
    axis.text.x = element_blank()) +
  scale_color_manual(values = c("cyan4", "orange3")) +
  theme(legend.position = c(0.9, 0.8)) +
  theme(legend.background = element_rect(fill="white", 
                                  size=0, linetype="solid"))
```

```{r}
# using patchwork to combine plots
(plot_h | plot_l | plot_w)
#(plot_w | plot_gl)
```

### Binary logistic regression

Performing a BLR for two models:

- **Model 1:** Log odds of plant type using plant height, canopy length, canopy width and green leaves

- **Model 2:** Log odds of plant type using plant height, canopy width and green leaves

```{r}
# storing functions of combinations of variables for each model
f1 <- species_name ~ height + length + width + green_lvs
f2 <- species_name ~ height + width + green_lvs

# performing and storing the BLR for each model
blr1 <- glm(formula = f1, data = palmetto, family = 'binomial')
blr2 <- glm(formula = f2, data = palmetto, family = 'binomial')

# examining the outputs
#summary(blr1)
#summary(blr2)

blr1_tidy <- tidy(blr1)
blr2_tidy <- tidy(blr2)

blr1_tidy %>% 
  kable(caption = "**Table 1:** Model 1 BLR results",
        col.names = c("Variable", "Coefficient", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE)

blr2_tidy %>% 
  kable(caption = "**Table 2:** Model 2 BLR results",
        col.names = c("Variable", "Coefficient", "Standard error", "Statistic", "p-value")) %>% 
  kable_styling(full_width = FALSE)
```

### AIC and BIC

```{r}
aic <- aictab(list(blr1, blr2))
# model1 = 5194
# model2 = 5987
```

**Model 1:** `r round(aic$AICc[1], 1)`

**Model 2:** `r round(aic$AICc[2], 1)`

The difference in AIC between the two models is large (`r round(aic$Delta_AICc[2], 1)`), showing that model 1 is significantly better fit than model 2.

```{r}
bic <- bictab(list(blr1, blr2))
# model1 = 5232
# model2 = 6017
```

**Model 1:** `r round(bic$BIC[1], 1)`

**Model 2:** `r round(bic$BIC[2], 1)`

The difference in BIC between the two models is large (`r round(bic$Delta_BIC[2], 1)`), showing that model 1 is significantly better fit than model 2.

### Ten fold cross validation 

```{r}
set.seed(218) # setting seed for reproducibility

tidyfold <- vfold_cv(palmetto, v = 10, repeats = 5)

# defining model type 
blr_model <- logistic_reg() %>% 
  set_engine('glm')

# set up workflow that bundles logistic model and formula

# running for formula 1
blr_tidy_wf_1 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f1)

blr_tidy_cv_f1 <- blr_tidy_wf_1 %>%
  fit_resamples(tidyfold)

# outputting metrics
cv_metrics_f1 <- collect_metrics(blr_tidy_cv_f1)

# running for formula 2
blr_tidy_wf_2 <- workflow() %>%
  add_model(blr_model) %>%
  add_formula(f2)

blr_tidy_cv_f2 <- blr_tidy_wf_2 %>%
  fit_resamples(tidyfold)

# outputting  metrics
cv_metrics_f2 <- collect_metrics(blr_tidy_cv_f2) 
 
# create tables for cross validation metrics

# model 1
cv_metrics_f1 %>% 
  kable(caption = '**Table 3:** Model 1 Cross Validation',
        col.names = c('Metric', 'Estimator', 'Mean',
                      'n', 'Standard error', 'Configuration')) %>% 
  kable_styling(full_width = FALSE)

# model 2
cv_metrics_f2 %>% 
  kable(caption = '**Table 4:** Model 2 Cross Validation',
        col.names = c('Metric', 'Estimator', 'Mean',
                      'n', 'Standard error', 'Configuration')) %>% 
  kable_styling(full_width = FALSE)
```

As shown in Tables 3 and 4, the mean accuracy of Model 1 is `r round(cv_metrics_f1$mean[1],3)`, while the mean accuracy for Model 2 is `r round(cv_metrics_f2$mean[1],3)`. These cross validation results, combined with the AIC and BIC results, show that Model 1 performs better than Model 2 at species classification.

### Training Model 1 on the entrie data set

```{r}
# using the entire data set to train model 1
blr1_tidyfit <- blr_model %>% 
  fit(f1, data = palmetto)

# output coefficients into a table
coef_blr1_tidy <- tidy(blr1_tidyfit)
coef_blr1_tidy %>% 
  kable(caption = '**Table 5:** Final Model BLR coefficients',
        col.names = c('Variable', 'Coefficient', 'Standard error', 'Statistic', 'p-value')) %>% 
  kable_styling(full_width = FALSE)
```

### Model accuracy

```{r}
# converting log odds to probabilities
blr_fitted <- blr1 %>% 
  broom::augment(type.predict = "response")

# creating a function to calculate accuracy
pred_acc <- function(x, y) {
  accurate <- ifelse(x == y, 0, 1)
  return(accurate)
}

model_success <- blr_fitted %>% 
  mutate(predict_repens = predict(blr1, palmetto, type = "response")) %>% 
  mutate(predicted = ifelse(predict_repens >= 0.50, "Serenoa repens", "Sabal etonia")) %>%
  mutate(accurate = pred_acc(species_name, predicted)) %>% 
  mutate(correct =  case_when(species_name == predicted ~ "Correct",
                              TRUE ~ "Incorrect"))

pred_table <- model_success %>%
  group_by(species_name) %>% 
  summarize(total_accurate = sum(accurate),
            total_inaccurate = n() - sum(accurate),
            model_success = mean(accurate)*100)

pred_table %>% 
  kbl(digits = 2, 
      caption = "**Table 6:** Model Accuracy by Species", 
      col.names = c("Species", "Total Accurate", "Total Inaccurate", "% Correctly Classified")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

